DP Problems of Interest

### 2012-12-30 docs/roadmap/draft.tex
Nits. Missing: Zucker, proper user function codegen, report.

\subsection*{Abstract}
Dynamic programming is a common pattern of Computer Science used in various domains. Yet underlying matrix recurrences might be difficult to express and error prone. Additionally, domain experts might not have the skills to make an efficient parallel implementation. In this project, we leverage the Scala and LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs. Our contributions are: \ul
\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
%\item Reuse of existing compiler technology (fusion) for a specific purpose
%\item State of the art parallel implementation of these classes on GPUs
%\item Normalization of the grammar into efficient productions
%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
\ule
\vfill
This project has been achieved in collaboration with Manohar Jonnalagedda. I would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions. I hope you will enjoy your reading. \vspace{.3cm}\\ 
\textit{Thierry Coppey}

>>>>

\subsection*{Abstract}
Dynamic programming is a common pattern of Computer Science used in various domains. Yet underlying matrix recurrences might be difficult to express and error prone. Additionally, domain experts might not have the skills to make an efficient parallel implementation. In this project, we leverage the Scala and LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs. Our contributions are: \ul
\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
%\item Reuse of existing compiler technology (fusion) for a specific purpose
%\item State of the art parallel implementation of these classes on GPUs
%\item Normalization of the grammar into efficient productions
%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
\ule
\vfill
This project has been achieved in collaboration with Manohar Jonnalagedda. I also would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions. I hope you will enjoy your reading. \vspace{.3cm}\\ 
\textit{Thierry Coppey}

### 2012-12-30 docs/roadmap/problems.tex
Nits. Missing: Zucker, proper user function codegen, report.

@@

% ----------------------------------------------
\newpage
{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}
\subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
\item Matrices: $M_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		0 & stop\\
		M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
		M_{(i-1,j)}-d & N\\
		M_{(i,j-1)}-d & W
	\end{array}\right\}=B_{(i,j)} \]

@@

\item Optimizations:\ul
	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built). This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
	\item Padding: since first line and column of the matrix are zeroes, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome. Also since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned. Hence valid input indices are $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$. We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
	\ule
\ole

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with affine gap extension cost}\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

That can be written alternatively as:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
\end{array}\right\}=B_{(i,j)} \]

@@

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap. Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}. Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
	\ule

>>>>

Although not necessary to understand the project, the description of some of the most common dynamic programming problem is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems. Would the reader be familiar with dynamic programming, he could immediately jump to the next section.

@@

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman (simple)}\label{sswat}\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
\item Matrices: $M_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		0 & stop\\
		M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
		M_{(i-1,j)}-d & N\\
		M_{(i,j-1)}-d & W
	\end{array}\right\}=B_{(i,j)} \]

@@

\item Optimizations:\ul
	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built). This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
	\item Since first line and column of the matrix (corresponding to a matching with one string being empty) are zeroed, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$ thus we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned. Hence valid input indices would become $S[1 \cdots m-1]$ and $T[1 \cdots n-1]$. We refer as such strings as padded strings hereafter (with $|S_{\rm padded}| = |S| + 1$).
	\ule
\ole

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with affine gap extension cost}\ol

\item Problem: matching two strings $S$, $T$ with $|S|=m-1, |T|=n-1$.
\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

That can be written alternatively as:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
\end{array}\right\}=B_{(i,j)} \]

@@

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with arbitrary gap cost}\label{aswat}

{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}

\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap. Without loss of generality, let $m\ge n$\footnote{Otherwise if $|T|>|N|$ we only need to swap both the inputs and backtracking pairs.}. Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching and smaller gaps.} $g(x)=m-x$.
\item Matrices: $M_{m \times n}, B_{m \times n \times m}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
	\ule
